---
title: "Tick abundance Analyisis"
author: "Gerardo Ramos"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
### Lets load our packages first
```{r load-packages, message=FALSE}
library(tidyverse)
library(tidyr)
library(ggplot2)
library(emmeans)
library(multcompView)
library(multcomp)
```

### set working directory for all chunks in this file (default working directory is wherever Rmd file is)
```{r, setup, include=FALSE}
setwd("~/Documents/GitHub/Lyme-disease-disk-in-ecotone-habitats/Dataset")
```

## Lets read in our data 
```{r}
ticks<-read.csv("Plot_Data copy.csv", header = TRUE, stringsAsFactors = TRUE)
```

# Null Hypothesis: There are equal abundances of *Ixodes pacificus* in forest interiors, forest edge, and grasslands

# Alternative Hypothesis: The abudances of *Ixodes pacificus* are different between forest interiors, forest edge, and grasslands



## Make a boxplot & violin plotand IPACs per habitat

```{r}
ticks$Habitat <- factor(ticks$Habitat, levels = c("exterior", "edge", "interior"))

 p1 <- ggplot(ticks,aes(x=Habitat,y=IPAC_N, fill = Habitat)) +
     geom_boxplot() +
     theme_classic(15) +
     ylab("Nymphal Ixodes pacificus Count") +
     xlab("Habitat Type")+
     scale_fill_manual(values = c("exterior" = "#FEF851", "edge" = "#A1B234", "interior" = "#4DA829"))+
   theme(
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.title = element_text(size = 20),  # Title of the legend
    axis.text.x = element_text(size = 20),  # x-axis category labels
   legend.text = element_text(size = 18)    # Labels inside the legend
  )
p1

 p2 <- ggplot(ticks,aes(x=Habitat,y=IPAC_L, fill = Habitat)) +
     geom_boxplot() +
     theme_classic(15) +
     ylab("Larval Ixodes pacificus Count") +
     xlab("Habitat Type")+
     scale_fill_manual(values = c("exterior" = "#FEF851", "edge" = "#A1B234", "interior" = "#4DA829"))+
   theme(
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.title = element_text(size = 20),  # Title of the legend
    axis.text.x = element_text(size = 20),  # x-axis category labels
   legend.text = element_text(size = 18)    # Labels inside the legend
  )
p2

 p3 <- ggplot(ticks,aes(x=Habitat,y=IPAC_A, fill = Habitat)) +
     geom_boxplot() +
     theme_classic(15) +
     ylab("Adult Ixodes pacificus Count") +
     xlab("Habitat Type")+
     scale_fill_manual(values = c("exterior" = "#FEF851", "edge" = "#A1B234", "interior" = "#4DA829"))+
   theme(
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.title = element_text(size = 20),  # Title of the legend
    axis.text.x = element_text(size = 20),  # x-axis category labels
   legend.text = element_text(size = 18)    # Labels inside the legend
  )
p3
```



## Make a boxplot & violin plot for DEOC Adults 
```{r}
 p4 <- ggplot(ticks,aes(x=Habitat,y=DEOC_A, fill = Habitat)) +
     geom_boxplot() +
     theme_classic(15) +
     ylab("Adult Ixodes pacificus Count") +
     xlab("Habitat Type")+
     scale_fill_manual(values = c("exterior" = "#FEF851", "edge" = "#A1B234", "interior" = "#4DA829"))+
   theme(
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.title = element_text(size = 20),  # Title of the legend
    axis.text.x = element_text(size = 20),  # x-axis category labels
   legend.text = element_text(size = 18)    # Labels inside the legend
  )
p4
```

## Lets look at the assumptions.
```{r}
hist(ticks$IPAC_L)
hist(ticks$IPAC_N)
hist(ticks$DEOC_A)
```

## All of these are skewed to the right and are count data so that means we can use a poisson family for our GLMM. 

## Lets start with the larvae
```{r}
m1<- glmmTMB(IPAC_L ~ Habitat +
                   (1 | Site),
                   family = poisson(link = "log"),
                   data = ticks)
m2<- glmmTMB(IPAC_L ~ Habitat +
                   (1 | Site) + (1 | Date),
                   family = poisson(link = "log"),
                   data = ticks)
m3<- glmmTMB(IPAC_L ~ Habitat,
                   family = poisson(link = "log"),
                   data = ticks)


AIC(m1, m2, m3)

summary(m2)

```


## Lets do IPAC Nymphs
```{r}
m4<- glmmTMB(IPAC_N ~ Habitat +
                   (1 | Site),
                   family = poisson(link = "log"),
                   data = ticks)
m5<- glmmTMB(IPAC_N ~ Habitat +
                   (1 | Site) + (1 | Date),
                   family = poisson(link = "log"),
                   data = ticks)
m6<- glmmTMB(IPAC_N ~ Habitat,
                   family = poisson(link = "log"),
                   data = ticks)


AIC(m4, m5, m6)

summary(m5)

```



## Lets do IPAC DEOC
```{r}
m7<- glmmTMB(DEOC_A ~ Habitat +
                   (1 | Site),
                   family = poisson(link = "log"),
                   data = ticks)
m8<- glmmTMB(DEOC_A ~ Habitat +
                   (1 | Site) + (1 | Date),
                   family = poisson(link = "log"),
                   data = ticks)
m9<- glmmTMB(DEOC_A ~ Habitat,
                   family = poisson(link = "log"),
                   data = ticks)


AIC(m7, m8, m9)

summary(m8)
```


## Now lets run the emmeans for the models that we picked. 

## Lets start with IPAC larvae
```{r}
library(emmeans)
em2 <- emmeans(m2, ~ Habitat)

# Pairwise comparisons between habitat types
pairs(em2, adjust = "tukey")
```
## Lets start with IPAC Nymphs
```{r}
library(emmeans)
em5 <- emmeans(m5, ~ Habitat)

# Pairwise comparisons between habitat types
pairs(em5, adjust = "tukey")
```


## Lets start with DEOC Adults
```{r}
library(emmeans)
em8 <- emmeans(m8, ~ Habitat)

# Pairwise comparisons between habitat types
pairs(em8, adjust = "tukey")
```


## Finalize Figures for IPAC N
```{r}
p1+ annotate("text", x = 1, y = 3, label = "A", size = 6, fontface = "bold") + annotate("text", x = 2, y = 22, label = "B", size = 6, fontface = "bold") + annotate("text", x = 3, y = 15, label = "C", size = 6, fontface = "bold")
```

## Finalize Figures for DEOC adults 
```{r}
p4+ annotate("text", x = 1, y = 6, label = "AB", size = 6, fontface = "bold") + annotate("text", x = 2, y = 7, label = "A", size = 6, fontface = "bold") + annotate("text", x = 3, y = 4, label = "B", size = 6, fontface = "bold")
```



# Results
For nymphal *Ixodes pacificus* we observed that their populations increase as you move along a gradient from grassland to oakwoodlands. Each habitat surveyed had a significanlty different abundance of ticks. For our *Dermocentor occidentalis* adults we observed the most abundances in grasslands (exterior) forest edge, while interior experienced the lowest abundances.